Fast and Flexible Binary Instrumentation for the Kernel

### 2013-12-05 VEE-14/paper.tex
Working on the first few comments on ashvin's second review of the paper.

\subsection{Basic Block Meta-Data}

@@

	\item[iii) Improved visibility.] Not ending basic blocks on \texttt{call} instructions results in longer basic blocks, which in turn provide additional visibility to tools at instrumentation time into what will be executed by the current thread. \toolname{RCUdbg} benefits from this additional visibility because it can observe the end of read-side critical sections (\texttt{call}s to \texttt{rcu\_read\_unlock}), and optimistically assume that functions executing before the read-side critical section ends will execute within the read-side critical section.

	\item[ii) Zero-overhead attaching.] (\Cref{sec:implicit_attach})

@@

@@

DRK's solution to this problem assumes that all injected instrumentation instructions are potentially non-reentrant, and \emph{delays} interrupts occuring within instrumented code \cite{DRK}. Interrupt delaying allows DRK to maintain the precise interrupt semantics of x86: interrupts only arrive on \emph{logical} instruction boundaries, and kernel interrupt handlers only observe faults or exceptions at native instruction boundaries. This approach introduces subtle complexity when creating instrumentation that injects instructions both before (pre) and after (post) native instructions. For example, an interrupt occuring in the pre-instrumentation of a native instruction will be delayed only until the native instruction. When the interrupt is handled and execution attempts to return to the interrupted instructions, DRK re-instruments the ``tail" of interrupted basic block, starting from the native instruction. Instrumenting this tail involves injecting pre- and post-instrumentation around the first native instruction. If the pre-instrumentation is not idempotent, then its re-execution will result in undefined behavior. We encountered this issue when using DRK for kernel module instrumentation, where our pre-instrumentation spilled registers to the runtime call stack, and our post-instrumentation restored the spilled registers. Because of interrupt delaying, the spilling code was sometimes repeated many times, whereas the restoring was only executed once. This led to hard-to-debug kernel crashes.

@@

Granary charts a middle ground between the two aforementioned approaches. Interrupt delaying has value because it enables non-reentrant instrumentation to be injected without costly disabling/enabling of interrupts, while not supporting delaying has value because not all instrumentation is non-reentrant. By default, Granary does not delay interrupts. However, instrumentation tools can declare \emph{interrupt delay ranges}: sequences of instructions (which can contain internal control flow) inside of a basic block that must be executed atomically. Granary encodes bounds information about interrupt delay ranges inside of each basic block's meta-data. When a basic block is interrupted, Granary inspects the meta-data and determines if the interrupt must be delayed. If a delay is required, then the instructions belonging to the delay range are copied into a CPU-private buffer, followed by code that re-constructs the interrupt stack frame and re-issues the interrupt to Granary. Granary then emulates a return from the current interrupt handler (without issuing an \texttt{iret} instruction), but leaves interrupts disabled and redirects execution into the copied code. When the copied code finishes, Granary regains control and observes the reconstructed interrupt stack frame. At this point, Granary or the tool can choose to handle the interrupt or defer the handling of the interrupt to the kernel.

@@

\subsection{Implementation}
\subsubsection{Direct Control Flow}
% Code block showing an example JZ (jump if ZF=1).
\lstset{language=[x64]Assembler}
\newsavebox\nativejcc
\begin{lrbox}{\nativejcc}
\begin{minipage}[b]{4cm}
\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
cmp $0x0, %rax
jz value_is_zero 
...
( fall through )
\end{lstlisting}
\end{minipage}
\end{lrbox}

@@

\begin{figure}[t!]
\subfloat[Native Code]{\usebox\nativejcc}
\hfill
\subfloat[Translated Code]{\usebox\translatedjcc}
\caption{\label{fig:direct_edge_code}Example translation of a conditional control flow instruction (\texttt{jz}) and its edge code. The original \texttt{jz} is translated into a \texttt{jz} that uses a 32-bit relative displacement to transfer control to an edge code function call (\texttt{edge\_value\_is\_zero}). When executed, the function call reads its target from an edge data structure (\texttt{jz\_patch\_data}), which contains all information necessary to resolve the \texttt{jz}'s target and patch the original \texttt{jz}. Patching is performed with a four byte atomic write to the relative displacement operand of the \texttt{jz} instruction. Finally, the return address of the \texttt{call} instruction is altered to transfer control to the newly resolved target.}
\end{figure}
Like btkernel, Granary translates direct control-flow instructions (CTIs) into equivalent jumps into CTI-specific ``edge code" (\Cref{fig:direct_edge_code}). Edge code lives outside the code cache, so as to not pollute the icache \cite{btkernel}.

When control flows through a basic block $B_{pred}$ for the first time, any jumps taken will transfer control to edge code that is specific to the jump taken. Edge code contains only a single indirect \texttt{call} instruction, thus allowing multiple edge codes to occupy a single cache line. The indirect \texttt{call} transfers control to Granary via a function pointer located in an ``edge data structure", which describes how to translate the target basic block, $B_{succ}$. Granary gains control and looks up or translates the target code according instrumentation policy information stored in the edge data structure. Then, Granary uses hot-code patching (\Cref{sec:patching}) to modify the original CTI that led to the edge code \texttt{call}, so that it now transfers control to $B_{succ}$ instead. Finally, Granary releases control and returns to $B_{succ}$.

@@

\subsection{Optimizations}
\subsubsection{Ahead-of-time Translation} \label{sec:aot}
Granary ends basic blocks at conditional branches (\texttt{jCC}), direct and indirect \texttt{jmp}s, and \texttt{ret}s and \texttt{iret}s. Basic blocks ending in conditional branches are extended to contain two trailing CTIs: a copy of the native \texttt{jCC}, and a synthesized direct \texttt{jmp} to the code executed when the condition code is not set (i.e. when the conditional jump is not taken).

@@

\subsubsection{Patching and Alignment}\label{sec:patching}

For efficiency, basic blocks in Granary's code cache are directly connected through direct control flow instructions (\texttt{call}, \texttt{jmp}, \texttt{jCC}). That is, if there is a direct branch from one native instruction to another, then there will be a corresponding direct branch between two translated instructions in Granary's code cache.

@@

%To reduce the overhead of alignment

%In some cases, basic blocks can be directly connected during basic block translation time. For example, intra-trace CTIs, or CTIs targeting already translated basic blocks. However, not all basic blocks can be connected this way, therefore some runtime

%Granary connects some basic blocks at translation time (e.g. intra-trace control-flow instructions). This is common when tracing is enabled, but also occurs when a branch targets an already translated basic block. However,

%Some basic blocks are directly connected at basic block translation time. One example is branches that target other basic blocks within a trace. Another example is a branch to an already translated basic block.

%Other basic blocks are only connected at runtime. For example, if the function \texttt{memset} has never been executed, then a \texttt{call} to \texttt{memset} will be translated to target edge code (\Cref{fig:direct_edge_code}) that, when executed, will translate \texttt{memset}'s first basic block and patch the \texttt{call} to target the translated basic block.

%For example, the targets of function call instruction (\texttt{call}) are resolved just-in-time (unless the target is already known). For each translated \texttt{call} instruction, Granary generates edge code (\Cref{fig:direct_edge_code}) that will find or translate the first basic block of the function being called, and patch the translated \texttt{call} instruction to point to the function's first basic block.

%For example, if a direct branch in one basic block targets another block within the trace under construction, then that

%Basic blocks are directly connected at block translation time when branches in traces re-enter a trace, or when branches target previously translated basic blocks. If a branch does not target another basic block within its trace, or does not target a previously translated basic block, then the branch will target edge code that will resolve that branches target at runtime.

% However, when a given basic block is translated, not all of its targets are known (e.g. \texttt{call} and \texttt{jCC} targets). Therefore, edge code (\Cref{fig:direct_edge_code}) is generated that translates/resolves a branch target when that branch is first executed, and patches the translated branch to avoi

% The targets of some of these direct CTIs are only resolved at runtime, by executing edge code. As an optimization, jumps to edge code are dynamically patched to transfer control to their intended targeted basic blocks.

%edge code use hot code patching to update a translated CTI to redirect execution

@@

%Directing control through the shadow system call table avoids the overhead of going through the indirect branch lookup (IBL) table (\Cref{sec:ibl}), and reduces the number of potential conflicts in the IBL table.

% whose entries are code cache basic block addresses translated according to the starting instrumentation policy.

%\subsection{Reproducability}
%Hello

@@

@@

While sufficient for comprehensiveness, hardware page protection is an inefficient mechanism for gaining control of module code execution. That is, after the (instrumented) module initializer runs to completion, Granary needs an efficient way to regain control of future invocations of the module's code. Our approach to regaining control is based on the observation that modules tell the kernel about their interfaces by registering functions with the kernel. We expect that at least some of the registered functions will be executed by the kernel because this is the mechanism by which modules extend the kernel's functionality. To Granary, registered module functions represent potential future \emph{attach} points, where a switch from native to instrumented execution will occur. Furthermore, Granary discovers additional attach points at detach points by observing pointers to module functions that are passed by modules to the kernel.

@@

\begin{figure}
\captionsetup[subfloat]{width=0.45\columnwidth}
\begin{tabularx}{\columnwidth}{XX}%
\subfloat[Full kernel instrumentation.]{\hspace{2.5em}\epsfig{file=diagrams/kernel.eps,width=0.125\textwidth}} & %
\subfloat[Module-only instrumentation]{\hspace{2.5em}\epsfig{file=diagrams/module.eps,width=0.125\textwidth}} \\
\subfloat[Module-only instrumentation, with a function specialized to run natively by using a module function wrapper.]{\hspace{2.5em}\epsfig{file=diagrams/module_wrapper.eps,width=0.125\textwidth}} & %
\subfloat[\label{fig:module_probe}Module-only instrumentation, with a kernel function specialized to also be instrumented by using a probe-based wrapper.]{\hspace{2.5em}\epsfig{file=diagrams/module_probe.eps,width=0.125\textwidth}}
\end{tabularx}
%
%\begin{multicols}{2}
%\subfloat[Full kernel instrumentation.]{\epsfig{file=diagrams/kernel.eps,width=0.125\textwidth}}
%\columnbreak
%\subfloat[Module-only instrumentation]{\epsfig{file=diagrams/module.eps,width=0.125\textwidth}}
%\end{multicols}
%\begin{multicols}{2}
%\subfloat[Module-only instrumentation, with a function specialized to run natively by using a module function wrapper.]{\epsfig{file=diagrams/module_wrapper.eps,width=0.125\textwidth}}
%\columnbreak
%\subfloat[Module-only instrumentation, with a kernel function specialized to also be instrumented by using a probe-based wrapper.]{\epsfig{file=diagrams/module_probe.eps,width=0.125\textwidth}}
%\end{multicols}
\caption{Example of using different granularities of instrumentation to selectively instrument or not instrument specific code.}
\end{figure}

>>>>

\subsection{Basic Block Meta-Data}\label{sec:metadata}

@@

	\item[ii) Improved visibility.] Not ending basic blocks on \texttt{call} instructions results in longer basic blocks, which in turn provide additional visibility to tools at instrumentation time into what will be executed by the current thread. \toolname{RCUdbg} benefits from this additional visibility because it can observe the end of read-side critical sections (\texttt{call}s to \texttt{rcu\_read\_unlock}), and optimistically assume that functions executing before the read-side critical section ends will execute within the read-side critical section.

@@

\begin{figure}[t!]
\begin{multicols}{2}
\vspace{-1.5em}\subfloat[\label{fig:interrupt_delay_drk}Pre and post instrumentation in DRK. Injected instrumentation blocks are treated as atomic with respect to interrupts.]{\includegraphics[height=10em,clip=true,trim=0 5em 10em 0]{diagrams/drk_instrumented_mov.pdf}}\vspace{-2.5em}
\columnbreak
\vspace{-1.5em}\hspace{-0.5em}\subfloat[\label{fig:interrupt_delay_granary}Pre and post instrumentation in Granary. The two instrumentation blocks, as well as the native instruction are wrapped in a delay region.]{\includegraphics[height=10em,clip=true,trim=0 5em 10em 0]{diagrams/granary_instrumented_mov.pdf}}\vspace{-2.5em}
\end{multicols}
\caption{Example showing equivalent pre- and post-instruction instrumentation with DRK (left) and Granary (right). In DRK, the injected instrumentation instructions (grey) are guaranteed to execute atomically, up until the next native instruction (white). In Granary, no guarantees are made on injected instrumentation, unless otherwise instructed. In this case, a delay region surrounds both the pre- and post-instrumentation, as well as the native instruction, guaranteeing that the entire sequence of instructions execute on the same CPU.}
\end{figure}

@@

DRK's solution to this problem assumes that all injected instrumentation instructions are potentially non-reentrant, and \emph{delays} interrupts occuring within instrumented code \cite{DRK}. Interrupt delaying allows DRK to maintain the precise interrupt semantics of x86: interrupts and exceptions only arrive on logical instruction boundaries in the native code, and kernel interrupt handlers only observe faults or exceptions at native instruction boundaries. This approach introduces subtle complexity when creating instrumentation that injects instructions both before (pre) and after (post) native instructions.

\Cref{fig:interrupt_delay_drk} shows an example of pre- and post-instrumentation around a \texttt{mov} instruction. If an interrupt arrives within the pre-instrumentation, then DRK delays they delivery of the interrupt until the native instruction (\texttt{mov}). After the interrupt is handled and execution attempts to return to the interrupted instructions, DRK will re-instrument the ``tail" of interrupted basic block, starting from the \texttt{mov} instruction. This involves re-creating the pre- and post-instrumentation of the \texttt{mov} in the tail basic block. When executed, this will result in an unbalanced number of \texttt{push} and \texttt{pop} instructions. That is, \texttt{push \%rax} is executed in both the original and tail basic block, but \texttt{pop \%rax} is only executed in the tail basic block. This is a specific example of a more general problem: DRK's approach to interrupt delaying does not play well with idempotent instrumentation.

%If the pre-instrumentation is not idempotent, then its re-execution will result in undefined behavior. We encountered this issue when using DRK for kernel module instrumentation, where our pre-instrumentation spilled registers to the runtime call stack, and our post-instrumentation restored the spilled registers. Because of interrupt delaying, the spilling code was sometimes repeated many times, whereas the restoring was only executed once. This led to hard-to-debug kernel crashes.

@@

Granary charts a middle ground between the two aforementioned approaches. Most instrumentation is re-entrant, partially because kernel instrumentation tools can depend on the runtime call stack for scratch storage space. However, interrupt delaying is sometimes useful because it enables non-reentrant instrumentation to avoid to continually disable and re-enable interrupts. For example, Granary's basic block tracing tool writes to a CPU-private buffer containing the sequence of executed basic block program counters. Operating on a CPU-private buffer is more efficient than operating on a global buffer because it avoids cache line contention and locking on an SMP. However, writing to a CPU-private buffer is only safe if the entire write executes on the same CPU. An equivalent tool developed for btkernel---which does not support interrupt delaying---would have to disable and the re-enable interrupts around the write. The is inefficient because it introduces flag stalls into the instruction pipeline \cite{AgnerMicroarchitecture}, and because interrupts arrive relatively infrequently.

Granary conveniently solves this problem by allowing instrumentation tools to declare \emph{interrupt delay regions}. An interrupt delay region is a sequence of instructions (which can contain internal control flow) inside of a basic block that must be executed atomically. \Cref{fig:interrupt_delay_granary} shows an interrut delay region that surrounds both the pre- and post-instrumentation of a native \texttt{mov} instruction. If an interrupt arrives anywhere within this region of code, Granary will delay the interrupt to the end of the region. The declarative nature of interrupt delay regions allows instrumentation developers to opt-in to interrupt delaying, and control exactly when, where, and for how long interrupts can be delayed.

Granary implements interrupt delay regions by recording bounds information about interrupt delay regions inside of each basic block's meta-data (\Cref{sec:metadata}). When a basic block is interrupted, Granary inspects the block's meta-data and determines if the interrupt must be delayed. If a delay is required, then all instructions belonging to the delay range are copied into a CPU-private buffer. All instructions are copied because a delay region might contain internal control-flow. A CPU-private buffer is used because the number of CPUs is a natural upper bound on the number of simultaneously executing delay regions. After the copied instructions, Granary emits code that code re-creates the interrupt stack frame, followed by a jump to the interrupt vector (which is controlled by Granary). This jump ensures that Granary's interrupt handler regains control after the instructions within the delay region fully execute, but before the kernel observes the interrupt. When Granary regains control, it can then delegate to the instrumentation tool to handle the interrupt, or defer handling of the interrupt to the kernel.

When all instructions have been emitted to the CPU-private buffer, Granary emulates a return from the interrupt (without issuing an \texttt{iret} instruction). The emulated return leaves interrupts disabled and transfers control to the CPU-private buffer. Leaving interrupts disabled ensures that the remainder of the delay region will execute on the current CPU, and avoids reentrancy issues related to the copied region being located in a CPU-private buffer.

% and re-issues the interrupt to Granary. Granary then emulates a return from the current interrupt handler (without issuing an \texttt{iret} instruction), but leaves interrupts disabled and redirects execution into the copied code.\footnote{Leaving interrupts disabled is required to ensure that executing the copied instructions in the CPU-private buffer is reentrant. Granary performs some sanity checks on the instructions being copied to ensure that they don't re-enable interrupts.} When the copied code finishes, Granary regains control and observes the reconstructed interrupt stack frame. At this point, Granary or the tool can choose to handle the interrupt or defer the handling of the interrupt to the kernel.

%while not supporting delaying has value because not all instrumentation is non-reentrant. By default, Granary does not delay interrupts. However, instrumentation tools can declare \emph{interrupt delay ranges}:

@@

\subsection{Control Flow Implementation}
\subsubsection{Direct Control Flow}
% Code block showing an example JZ (jump if ZF=1).
\lstset{language=[x64]Assembler}
\newsavebox\nativejcc
\begin{lrbox}{\nativejcc}
\begin{minipage}[b]{4cm}
\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
cmp $0x0, %rax
jz value_is_zero 
...
( fall through )
\end{lstlisting}
\end{minipage}
\end{lrbox}

@@

\begin{figure}[t!]
\subfloat[Native Code]{\usebox\nativejcc}
\hfill
\subfloat[Translated Code]{\usebox\translatedjcc}
\caption{\label{fig:direct_edge_code}Example translation of a conditional control flow instruction (\texttt{jz}) and its edge code. The original \texttt{jz} is translated into a \texttt{jz} that uses a 32-bit relative displacement to transfer control to an edge code function call (\texttt{edge\_value\_is\_zero}). When executed, the function call reads its target from an edge data structure (\texttt{jz\_patch\_data}), which contains all information necessary to resolve the \texttt{jz}'s target and patch the original \texttt{jz}. Patching is performed with a four byte atomic write to the relative displacement operand of the \texttt{jz} instruction. Finally, the return address of the \texttt{call} instruction is altered to transfer control to the newly resolved target.}
\end{figure}
Like btkernel, Granary translates direct control-flow instructions (CTIs) into equivalent jumps into CTI-specific ``edge code" (\Cref{fig:direct_edge_code}). Edge code lives outside the code cache, so as to not pollute the processor's instruction cache \cite{btkernel}.

When control flows through a basic block $B_{pred}$ for the first time, any jumps taken will transfer control to edge code that is specific to the jump taken. Edge code contains only a single indirect \texttt{call} instruction, thus allowing multiple edge codes to occupy a single cache line. The indirect \texttt{call} transfers control to Granary via a function pointer located in an ``edge data structure", which describes how to translate the target basic block, $B_{succ}$. Granary gains control and looks up or translates the target code according instrumentation policy stored in the edge data structure. Then, Granary uses modifies the original CTI in $B_{pred}$ in place so that it transfers control to $B_{succ}$  (\Cref{sec:patching}). Finally, Granary releases control and context switches to $B_{succ}$.

@@

\subsection{Optimizations}

\subsubsection{Ahead-of-time Translation} \label{sec:aot}
Granary ends basic blocks at conditional branches (\texttt{jCC}), direct and indirect \texttt{jmp}s, and \texttt{ret}s and \texttt{iret}s. Basic blocks ending in conditional branches are extended to contain two trailing CTIs: a copy of the native \texttt{jCC}, and a synthesized direct \texttt{jmp} to the code executed when the condition code is not set (i.e. when the conditional jump is not taken).

@@

\subsubsection{Patching and Alignment}\label{sec:patching}
For efficiency, basic blocks in Granary's code cache are directly connected through direct control flow instructions (\texttt{call}, \texttt{jmp}, \texttt{jCC}). That is, if there is a direct branch from one native instruction to another, then there will be a corresponding direct branch between two translated instructions in Granary's code cache.

@@

%To reduce the overhead of alignment 
%In some cases, basic blocks can be directly connected during basic block translation time. For example, intra-trace CTIs, or CTIs targeting already translated basic blocks. However, not all basic blocks can be connected this way, therefore some runtime 
%Granary connects some basic blocks at translation time (e.g. intra-trace control-flow instructions). This is common when tracing is enabled, but also occurs when a branch targets an already translated basic block. However, 
%Some basic blocks are directly connected at basic block translation time. One example is branches that target other basic blocks within a trace. Another example is a branch to an already translated basic block.
%Other basic blocks are only connected at runtime. For example, if the function \texttt{memset} has never been executed, then a \texttt{call} to \texttt{memset} will be translated to target edge code (\Cref{fig:direct_edge_code}) that, when executed, will translate \texttt{memset}'s first basic block and patch the \texttt{call} to target the translated basic block.
%For example, the targets of function call instruction (\texttt{call}) are resolved just-in-time (unless the target is already known). For each translated \texttt{call} instruction, Granary generates edge code (\Cref{fig:direct_edge_code}) that will find or translate the first basic block of the function being called, and patch the translated \texttt{call} instruction to point to the function's first basic block.
%For example, if a direct branch in one basic block targets another block within the trace under construction, then that 
%Basic blocks are directly connected at block translation time when branches in traces re-enter a trace, or when branches target previously translated basic blocks. If a branch does not target another basic block within its trace, or does not target a previously translated basic block, then the branch will target edge code that will resolve that branches target at runtime.
% However, when a given basic block is translated, not all of its targets are known (e.g. \texttt{call} and \texttt{jCC} targets). Therefore, edge code (\Cref{fig:direct_edge_code}) is generated that translates/resolves a branch target when that branch is first executed, and patches the translated branch to avoi
% The targets of some of these direct CTIs are only resolved at runtime, by executing edge code. As an optimization, jumps to edge code are dynamically patched to transfer control to their intended targeted basic blocks.
%edge code use hot code patching to update a translated CTI to redirect execution

@@

%Directing control through the shadow system call table avoids the overhead of going through the indirect branch lookup (IBL) table (\Cref{sec:ibl}), and reduces the number of potential conflicts in the IBL table.
% whose entries are code cache basic block addresses translated according to the starting instrumentation policy.
%\subsection{Reproducability}
%Hello

@@

\begin{figure}
\captionsetup[subfloat]{width=0.45\columnwidth}
\begin{tabularx}{\columnwidth}{XX}%
\subfloat[Full kernel instrumentation.]{\hspace{2.5em}\epsfig{file=diagrams/kernel.eps,width=0.125\textwidth}} & %
\subfloat[Module-only instrumentation]{\hspace{2.5em}\epsfig{file=diagrams/module.eps,width=0.125\textwidth}} \\
\subfloat[Module-only instrumentation, with a function specialized to run natively by using a module function wrapper.]{\hspace{2.5em}\epsfig{file=diagrams/module_wrapper.eps,width=0.125\textwidth}} & %
\subfloat[\label{fig:module_probe}Module-only instrumentation, with a kernel function specialized to also be instrumented by using a probe-based wrapper.]{\hspace{2.5em}\epsfig{file=diagrams/module_probe.eps,width=0.125\textwidth}}
\end{tabularx}
\caption{Example of using different granularities of instrumentation to selectively instrument or not instrument specific code.}
\end{figure}

@@

While correct, hardware page protection is an inefficient mechanism for gaining control of module code execution. That is, after the (instrumented) module initializer runs to completion, Granary needs an efficient way to regain control of future invocations of the module's code. Our approach to regaining control is based on the observation that modules tell the kernel about their interfaces by registering functions with the kernel. We expect that at least some of the registered functions will be executed by the kernel because this is the mechanism by which modules extend the kernel's functionality. To Granary, registered module functions represent potential future \emph{attach} points, where a switch from native to instrumented execution will occur. Furthermore, Granary discovers additional attach points at detach points by observing pointers to module functions that are passed by modules to the kernel.

@@
